# Cryonic Demography

```{r setup, include=FALSE}
base64stringify <- function(path){
  paste0("data:image/png;base64 ", knitr::image_uri(path))
}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  error = FALSE,
  fig.process=base64stringify
)
```

## Introduction

There's a lot of data available about patients in Cryonic suspension. For some public figures[e.g. Baseball star Ted Williams, Computer Scientist Hal Finney](sidenote), we know their identities and accordingly a great deal of their biographies. For others, we may simply know that a cryosuspension was attempted, with literally no other information about the patient.

The first cryonics patient falls quite close to the no-information end of the spectrum. We know that in April of 1966, an unidentified middle-aged woman died and was enbalmed and stored at near-freezing temperatures for two months, before being straight-frozen in liquid nitrogen by the Cryo-Care Equipment Corporation of Phoenix, Arizona. Within a year she was thawed.[For additional details, see <a href="https://www.alcor.org/library/suspension-failures-lessons-from-the-early-years/">M Perry. "Suspension Failures: Lessons from the Early Years." <i>Cryonics</i>. February 1992</a>](sidenote)

Cryonics doesn't have an especially long history or many patients, making it tractable to collect similar data on nearly all of them. Moreover, people in the cryonics community tend to be inclined towards more and better documentation--after all, who knows what data will be important at revival time? Also being of a generally transparent disposition (in the interest of de-mystifying the process for people who might be interested if the process and organization is laid bare before them), they tend to publish case information online:

* [Alcor](http://alcor.org/cases.html)
* [Cryonics Institute](http://www.cryonics.org/case-reports/)
* [KrioRus](http://kriorus.ru/en/cryopreserved%20people)
* [Oregon Cryonics](http://www.oregoncryo.com/caseReports.html)

We can collect these reports to begin to build a demographic profile of cyrosuspended patients.

## Collecting Data

Assembling all of the data isn't a trivial task. Many cryonics patients (or their survivors) prefer not to have their information shared publicly. Accordingly, the existing cryonics organizations' public records are spotty. The existing organizations are also not the only one to have ever existed, nor are they the first. Finding records from defunct organizations imposes a deep degree of detection, inference, and guess work (not to mention [earlier work](http://www.alcor.org/cryonics/cryonics9207.txt) and luck).[Since I first wrote this, another project led by Mati Roy has endeavoured to collect a wider swath of data on the existing cryonics patient population. <a href="https://docs.google.com/spreadsheets/d/1vBrlNvPkPQbLKxtSUlTF_ZpUS4lwW2W7lENRj9Gdk3I/edit?usp=sharing">See his spreadsheet here.</a>](sidenote)

```{r loadData, message=FALSE, warning=FALSE}
# Load from Google Sheets
library(readr)
library(dplyr)
patients <- read_csv('https://docs.google.com/spreadsheets/d/e/2PACX-1vS7nOdzuAUs5bSoHkZw1gJTWJT-MVbMicXK8iIUIkZwFqUVh2zmC8Cl_QTz1j0kriMW3ycobYDDWFuG/pub?output=csv') %>%
  mutate(DorCDate = as.Date(DorCDate))
dated <- patients %>% 
  dplyr::filter(!is.na(DorCDate))
```

### Managing Dates

Perhaps the most important data management decision for analysis regards the handling of dates. KrioRus reports "cryopreservation date", rather than the date of death[It should be noted that any mentions of "death" in this research should be interpreted as <i>legal</i> death, rather than <a href="https://en.wikipedia.org/wiki/Information-theoretic_death">information-theoretic death</a>](sidenote). However, it's a bit difficult to tell what this means, since cryopreservation is a process which takes place over the course of several days. The other organizations all report the date of death, except for [TransTime](https://www.transtime.com/) which reports nothing and whose cases must be inferred from other sources. Accordingly, where we have dates of death, we use those, and alternately use dates of cryopreservation.[This is referred to in the accompanying source code as the DorCDate, meaning "Death or Cryopreservation Date"](sidenote)

In practice, this imprecision shouldn't be too analytically meaningful, as the goal of this study is to profile demographics of cryonics patients over time, so most values will be aggregated up to the month. If a date disagreement should cause a patient to straddle the changes between months, then their record may be classified as the month before or after their actual legal death or cryopreservation.

## Population Growth

Cryonics, by virtue of beginning recently in history, has necessarily grown. But the dynamics of that growth are largely unexplored. To begin to peel back this mystery, the following graph shows the total number of people cryopreserved over time.

```{r growthOverTime, message=FALSE, warning=FALSE}
library(ggplot2)
cumulative <- dated %>%
  mutate(cumulation = 1:nrow(.))

ggplot(cumulative, aes(x=DorCDate, y=cumulation)) +
  geom_line() +
  labs(x = "Date", y = "Cryopreserved Patients")
```

Here we can clearly see the tell-tale ["hockey-stick" shape associated with exponential growth](https://www.oxfordreference.com/view/10.1093/oi/authority.20110803095939885). Let's leverage this to predict the future growth of the cryonically preserved population.

### Modeling Growth

I'm going to use a couple of models: linear (for posterity), quadratic (since it's the most intuitive curve that might fit this well), exponential (because this *actually* looks exponential). Given these three models, we can also create a simple "stacked" model--a linear model given the predictions of the others[Essentially, we can think of this as a way to compute the optimal weights for a weighted average of the model outputs.](sidenote). The graph below shows these models with the empirical data superimposed in black.

```{r growthModels, message=FALSE, warning=FALSE}
library(lubridate)
library(tidyr)
Year <- rep(min(year(patients$DorCDate), na.rm = TRUE):max(year(patients$DorCDate), na.rm = TRUE), each=12)
Month <- rep(1:12, length(Year)/12)
yms <- data.frame(Year, Month)

monthly <- dated %>%
  mutate(
    Year = year(DorCDate),
    Month = month(DorCDate)
  ) %>%
  group_by(Year, Month) %>%
  dplyr::summarise(number = n()) %>%
  ungroup() %>%
  right_join(yms) %>%
  dplyr::filter(!(Year >= year(now()) & Month >= month(now()))) %>%
  dplyr::arrange(Year, Month) %>%
  mutate(
    number = ifelse(is.na(number), 0, number),
    cumulation = cumsum(number),
    logCum = log1p(cumulation),
    YM = Year + (Month-1)/12,
    YM2 = YM^2,
    date = as.Date(paste(Year, Month, "01", sep = "-"))
  )

LinM  <- lm(cumulation ~ YM, data = monthly)
LinP  <- LinM %>% predict %>% unname
monthly <- cbind(monthly, LinP)

QuadM <- lm(cumulation ~ YM + YM2, data = monthly)
QuadP <- QuadM %>% predict %>% unname
monthly <- cbind(monthly, QuadP)

ExpM  <- lm(logCum ~ YM, data = monthly)
ExpP  <- ExpM %>% predict %>% unname %>% expm1
monthly <- cbind(monthly, ExpP)

AvgM  <- lm(cumulation ~ LinP + QuadP + ExpP, data = cbind(monthly, LinP, QuadP, ExpP))
AvgP  <- AvgM %>% predict %>% unname
monthly <- cbind(monthly, AvgP)

long <- monthly %>%
  pivot_longer(cols = c(LinP:AvgP), names_to="model")

ggplot(long, aes(x=date, y=value, color=model)) +
  geom_line() +
  geom_line(aes(y=cumulation), color="black") +
  labs(x = "Date", y = "Cryopreserved Patients")
```

Just by sight we can determine that the linear model (in blue) performs (expectedly) poorly, while the others appear to predict reasonably well. It's worth noting that near the end of the available data (i.e. more recently), the quadratic model (in purple) systematically underestimates the total while the exponential model (in green) systematically overestimates it. This is a convenient juxtaposition for the ensemble, as it means that an average of the two will necessarily be more accurate. Accordingly, we see that's the case--the ensemble (in orange) follows the empirical curve *very* closely. To suss out "how" exponential versus quadratic the phenomenon is, we need only consult the coefficients of the ensemble.

```{r}
library(broom)
library(knitr)
AvgM %>% tidy %>% kable(digits = 3)
```

As we can see, the linear model contributes very little signal to the overall model. Growth in the population of cryopreserved patients is non-linear. The quadratic model performs somewhat better, providing something approximating 25% of the variance[I know this is not how variance works in a strict statistical sense. Chill out.](sidenote). However, the core of the model is decidedly the exponential component, which accounts for more like 63% of the variance. In short, we can confidently claim that the population is undergoing exponential growth.

Given this confidence, what claims can we make about the course of the future of the cryopreserved population?

```{r cryonicsGrowthPrediction}
Year <- rep(min(year(patients$DorCDate), na.rm = TRUE):(max(year(patients$DorCDate), na.rm = TRUE)+40), each=12)
Month <- rep(1:12, length(Year)/12)
Fmonthly <- data.frame(Year, Month) %>%
  mutate(
    YM = Year + (Month-1)/12,
    YM2 = YM^2,
    date = as.Date(paste(Year, Month, "01", sep = "-"))
  )

FLinP  <- LinM %>% predict(Fmonthly) %>% unname
Fmonthly <- cbind(Fmonthly, LinP = FLinP)

FQuadP <- QuadM %>% predict(Fmonthly) %>% unname
Fmonthly <- cbind(Fmonthly, QuadP = FQuadP)

FExpP  <- ExpM %>% predict(Fmonthly) %>% unname %>% expm1
Fmonthly <- cbind(Fmonthly, ExpP = FExpP)

FAvgP  <- AvgM %>% predict(cbind(Fmonthly, LinP=FLinP, QuadP=FQuadP, ExpP=FExpP)) %>% unname
Fmonthly <- cbind(Fmonthly, AvgP = FAvgP)

long <- monthly %>%
  select(names(Fmonthly)) %>%
  rbind(Fmonthly) %>%
  mutate(cumulation = c(monthly$cumulation, rep(NA, nrow(Fmonthly)))) %>%
  pivot_longer(cols = c(LinP:AvgP), names_to="model")

ggplot(long, aes(x=date, y=value, color=model)) +
  geom_line() + 
  labs(x = "Date", y = "Cryopreserved Patients")
```

Running those same models into the future gives us a set of forecasts. As we can plainly see, the linear model (again, in blue) continues to woefully undershoot the others. After a point, it becomes clear that quadratic growth undershoots as well. This leaves us with the exponential model, which continues its optimistic course, and the ensemble, which follows a similar if slightly lower trajectory.

Taking the ensemble as the all-else-equal likeliest case, we should expect the cryopreserved population to continue its exponential growth, from mere hundreds today to nearly a thousand by 2030, and five thousand by 2055[Obviously, this discounts any substantial events, like a successful demonstration of a complete revival of a cryopreserved large animal](sidenote).

## Age and Gender

The simplest demographic features available within our data are age (at time of death) and gender. To start with, let's look at gender. It's been reported that [85% of cryonics signups are men](http://www.ozy.com/rising-stars/fighting-for-gender-equality-in-the-afterlife/73801).

```{r GenderOverTime, warning=FALSE}
w <- dated %>%
  group_by(Gender) %>%
  dplyr::mutate(
    c = 1,
    cumulation = cumsum(c)
  )

ggplot(w, aes(x = DorCDate, y = cumulation, color = Gender)) +
  geom_line() +
  labs(x = "Date", y = "Cryopreserved Patients")
```
Thus far, it seems men outnumber women roughly 3:2. However, the ratio was roughly 1:1 until the mid-90's. This is much closer to gender parity than I expected.

What about the distribution of age? This should be a left-skewed distribution (older people are more likely to die than younger people). Moreover, We can weakly predict that women will enjoy a slightly more left-skewed curve, outliving men in the cryonics community just as they tend to in the larger community.

```{r AgeDistributionByGender, warning=FALSE}
ggplot(patients, aes(Age, fill=Gender)) +
  geom_density(alpha = 0.5)
```

Unsurprisingly, the people who are cryopreserved are largely older. And it does look as though women outlive men by around 7 years in the cryonics community.

Does this match the curve of ages-at-death in the larger population? This is a surprisingly difficult question to answer! To start, let's look at the mean age-at-death of cryonics patients (along with the min and max ages, just for fun).

```{r AgeTrends, warning=FALSE}
ages <- patients %>%
  dplyr::filter(!is.na(DorCDate), !is.na(Age)) %>%
  mutate(
    MaxAge = cummax(Age),
    MeanAge = cummean(Age),
    MinAge = cummin(Age)
  ) %>%
  pivot_longer(MaxAge:MinAge, names_to = "Ages")

ggplot(ages, aes(x = DorCDate, y = value, color = Ages)) +
  geom_line() +
  labs(x = "Date", y = "Age")
```

The minimum age drops precipitously. Intuitively, this is about what we should expect: we can't cryopreserve people who aren't born yet, so the minimum age is going to hit zero at some point in the future, and never decrease again.

The maximum age climbs, which is also to be expected. As more people die and enter cryostorage, some of them should, just by random draws from the [Gompertz distribution](https://en.wikipedia.org/wiki/Gompertz_distribution), be older than the rest of the population. This should behave roughly like the curve of the record for the oldest person, gradually pushing up towards 120 or so (barring any major breakthroughs in longevity-extension efforts).

The mean population is the most interesting (and relevant to the motivating question). We expect that it should roughly trace the [global life expectancy](https://data.worldbank.org/indicator/SP.DYN.LE00.IN). However, because life expectancies are measured by country by year-of-birth, mapping from age at death (the data we have) to life expectancy (the data we need) is prohibitively difficult--we don't generally know the patients' birth country for one.

We can however, spot-check the mean age at death. The mean age of death has held steadily around 65 since the 2000's. These people would have been born around 1950. According to [Our World In Data](https://ourworldindata.org/life-expectancy/), the world-wide life expectancy in 1950 was 48, and the industrialized world (US, Europe, USSR, and Oceania) was closer to 60. So people planning to be cryopreserved are probably living at least a little longer than their generational cohort, on average. This may be a function of wealth, among other things: access to cryonics is expensive, and far from a priority budget item for most humans.

## Procedural Preferences

Not all cryonic procedures are exactly the same, and no cryonics provider offers all options. In general, there are two different types: whole body, which is basically exactly what you think, and neuropreservation, which is cryopreservation of only the head. More recently, Oregon Cryonics has begun performing brain-only preservations. and finally, there are separate whole body and neuro preservations.

```{r ProcedureDistributionOverTime, warning=FALSE}
procedures <- patients %>%
  dplyr::filter(!is.na(DorCDate), !is.na(OriginalType)) %>%
  group_by(OriginalType) %>%
  dplyr::mutate(
    c = 1,
    cumulation = cumsum(c)
  )

ggplot(procedures, aes(x = DorCDate, y = cumulation, color = OriginalType)) +
  geom_line() +
  labs(x = "Date", y = "Number of Cryonics Patients")
```

## Suspension Failure

Perhaps the most important and interesting inferences we can make regard the likelihood that a suspended person will be brought back. Because this is completely unknown (until someone is successfully revived), we can only make inferences based on estimates of probabilities of the things that may prevent someone from being revived. And the one bottleneck about which we have data is suspension failure.

```{r preservationFailures, warning=FALSE}
failures <- patients %>%
  dplyr::filter(!is.na(DorCDate), Status != "Preserved") %>%
  mutate(
    c = 1,
    cumulation = cumsum(c)
  )

ggplot(failures, aes(x = DorCDate, y = cumulation)) +
  geom_line() +
  labs(x = "Date", y = "Failed Suspensions")
```

Here we see the number of suspension failures over time.[Note that I didn't collect data on the date of the failure, so these are plotted against the date of death or cryopreservation](sidenote) Fortunately, there have been few enough that it almost makes the most sense to discuss them individually, rather than measure any aggregate statistics.

Of the 29 known suspension failures, the lion's share belong to the Cryonics Society of California (CSC). Led by the bold and hard-working but ill-fated Bob Nelson, CSC stored 10 patients, of whom all but one were lost in the [Chatsworth Disaster](https://hpluspedia.org/wiki/History_of_cryonics#Chatsworth_Scandal), hence the precipitous rise in the late 60's. Several others were cases of survivors refusing to make maintenance payments. Sadly, only one cryonics patient (Robert Ettinger) preserved before 1974 has been maintained. [Two were killed in the 9-11 World Trade Center attacks](http://alcor.org/problems.html), after which their remains could not be collected for preservation. [Two others (the Martinots) were private cases that suffered a freezer failure](https://www.theguardian.com/science/2006/mar/17/france.internationalnews). This leaves roughly half of the cryonics failures unaccounted for.

Relative to the number of successful cases, such failures are looking less and less likely:

```{r SuccessesAndFailures}
statuses <- dated %>%
  group_by(Status) %>%
  dplyr::mutate(
    c = 1,
    cumulation = cumsum(c)
  )

ggplot(statuses, aes(x = DorCDate, y = cumulation, color = Status)) +
  geom_line() +
  labs(x = "Date", y = "Cryonics Patients")
```

As time wears on, the proportion of failures is decreasing. So long as there aren't any facilities crises, it looks to continue this way. As a proportion of current cryonics cases, this failure rate may be a reasonable estimate for probabilistic cryonics failures in the future. So, what *is* that proportion?

```{r preservationProportion}
failures <- patients %>%
  dplyr::filter(!is.na(DorCDate)) %>%
  mutate(
    c = 1,
    failure = ifelse(Status=="Preserved", 0, 1),
    failures = cumsum(failure),
    proportion = 1-(failures / cumsum(c))
  )

ggplot(failures, aes(x = DorCDate, y = proportion)) +
  geom_line() +
  labs(x = "Date", y = "Proportion of Patients Still Cryopreserved")
```

94% and rising. This may form a useful piece of evidence for estimating the more general probability of cryonics succeeding. Here are some other attempts at estimating that probability:

* http://alcor.org/Library/html/WillCryonicsWork.html
* http://lesswrong.com/lw/7sj/how_likely_is_cryonics_to_work/
* http://www.merkle.com/cryo/probability.html
* https://www.jefftk.com/p/more-cryonics-probability-estimates
* https://www.gwern.net/Cryonics
* https://www.cryonicscalculator.com/

## Organizational History

What about the provider organizations?

```{r originalorganization}
providers <- patients %>%
  dplyr::filter(!is.na(DorCDate), !is.na(OriginalProvider)) %>%
  group_by(OriginalProvider) %>%
  dplyr::mutate(
    c = 1,
    cumulation = cumsum(c)
  )

ggplot(providers, aes(x = DorCDate, y = cumulation, color=OriginalProvider)) +
  geom_line() +
  labs(x = "Date", y = "Cryonics Patients")
```
[For comparison, Futurism put together an infographic including a graph of the growth of cryonics that largely agrees with this (though their version looked only at Alcor, Cryonics Institute and KrioRus statistics instead of all recorded cases, leading to slightly lower counts). <a href="https://futurism.com/images/cryonics/">Click here to see their version.</a>](marginnote)

## Future Work

There are tons more interesting things that could be done here, but the ones that interest me require additional data collection:

* Cause of death is known for many cases, even if I haven't recorded it. This may be a useful or interesting field for further investigation, especially as patients begin to be revived.
* Conditional on revivals happening, details on their ischemia periods and other post-mortem metrics may also be interesting.
* I haven't made any attempt to geocode the location data, but perhaps some maps would be interesting.

## Sources

* http://www.alcor.org/cryonics/cryonics9207.txt
* https://www.thisamericanlife.org/radio-archives/episode/354/mistakes-were-made/
* http://www.alcor.org/Library/html/suspensionfailures.html
* http://alcor.org/cases.html
* http://www.cryonics.org/case-reports/
* http://kriorus.ru/en/cryopreserved%20people
* http://www.oregoncryo.com/caseReports.html
* https://hpluspedia.org/wiki/Cryonics_patients
* http://www.cryonet.org/cgi-bin/dsp.cgi?msg=3247
* https://www.benbest.com/cryonics/history.html
* http://www.cryonics.org/ci-landing/patient-details/
* https://en.wikipedia.org/wiki/Category:Cryonically_preserved_people
